Higher order unification for free!

### 2024-04-09 paper/main.tex
more work

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is that the unification procedure \Ue of the meta language is not aware
of the equational theory of the object logic, even if both theories
include $\eta\beta$-conversion and admit most general
unifiers for unification problems in the pattern-fragment.

\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

\cref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution. \cref{sec:encodings} discusses alternative
term encodings and related works. \cref{sec:lang-spec} introduces
the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
basic simulation of higher order logic programs.
\cref{sec:eta,sec:invariant1} completes its equational theory
with support for $\eta$-conversion. \cref{sec:beta} deals with the
practical necessity of ``tolerating'' terms outside of the
pattern-fragment and discusses how heuristic can be applied.
Finally \cref{sec:implementation} discusses the implementation in Elpi.

@@

\newcommand{\specunif}[3]{
  #3_i \in \llambda \Rightarrow %
    \exists \rho, %
      \rho #3_1 #1 \rho #3_2  %
        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho
}

\newcommand{\unifcorrect}[3]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow
          \rho #3_1 #1 \rho #3_2  %
}

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #3_i \in \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
\newcommand{\notllambda}{\ensuremath{\overline{\llambda}}\xspace}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

The equational theory of Coq's Dependent Type Theory is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment \llambda.
We call this unification procedure \Uo{}.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:
%
\setlength{\tabcolsep}{2pt}
\vspace{4pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
\vspace{4pt}

To state precisely the problem we solve we need a \Fo{} representation
of CIC terms and a \Ho one. We extend the equality over ground
terms to open ones with the reflexivity of unification variables (that is
unification variable X is equal to itself, not to another one such a Y).
We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
We call \Uo the unification procedure we want to implement and
\Ue the one provided by the meta language.
We assume that the unification of our meta
language is correct and complete in \llambda.
%
%\begin{gather}
%  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
%  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
%\end{gather}

We write $t_1 \Ue t_2 \mapsto \sigma$ when $t_1$ and $t_2$ unify
with substitution $\sigma$; we write $\sigma t$ for the application of
the substitution to $t$, and $\sigma X = \{ \sigma t ~|~ t \in X\}$ when
$X$ is a set; we write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$ and we use set union to denote the concatenation of
two substitutions. We shall use $\rho$ for \Fo{} substitutions, and
$\sigma$ for the \Ho ones.

\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}

We illustrate a compilation $\C{s}{t}{m}{l}$ that
maps a term $s$ in \Fo{} to a term $t$ in \Ho{} and produces
a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$.
Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} $p$ of length $\mathcal{N}$.
Each step $p$ represents 
a unification problem between two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$.
\footnote{If the same rule is used multiple time in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time}
%
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
Here \hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links.
We claim:

@@

We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
%$\foUnifPb = \{ s_1, s_2 \}$ 
as follows:
%
$$
\begin{array}{l}
s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
\quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
\quad    t_1 \Ue t_2 \mapsto \sigma' \land
    \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
\quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
\end{array}
$$

\begin{proposition}[Properties of \Uo{} in \llambda]
\begin{gather}
  \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
  \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
\end{gather}
\end{proposition}

\noindent
In addition to correctness and completeness in \llambda, we claim that \hrun
handles terms outside \llambda in the following sense:

\begin{proposition}[Properties of \hrun outside \llambda]
\begin{gather}
  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow
  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \mapsto \sigma \label{prop:step-nllam}
\end{gather}
\end{proposition}

\noindent
Property \ref{prop:compilation-nllam} states that two terms for which
there is a unifier (given by an oracle, not necessarily a most general one),
then the compiler generates two terms that unify in \Ho. In other words
the compiler moves out of the way all problematic terms.

Property \ref{prop:step-nllam} says that if the two terms involved in a step
re-enter \llambda, then \hstep succeeds. This is a typical example
in which the order of the unification problems in a logic program run
do matter. The simplest example is the sequence $F = \lambda x.a$ and
$F \appsep a = a$: the second problem is not in \llambda and has two unifiers,
namely $F = \lambda x.x$ and $F = \lambda x.a$. The first problem picks one
of the two, making the second problem re-enter \llambda.

\subsection{The intuition in a nutshell}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
and an accessory \emph{link} that represent a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, that is it does not
contradict \Eo as it would otherwise do on ``problematic'' terms.
We now define ``problematic'' and ``well behaved'' more formally.

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n \not\in \llambda \}$.
\end{definition}

@@

Note that proposition \ref{prop:nf} does not hold for \Uo{} as a whole
since decompilation can introduce (actually restore) terms in
\maybeeta, \maybebeta or \notllambda that were moved out of the way
during compilation.

\section{Preliminaries: \Fo{} and \Ho}
\label{sec:lang-spec}

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax. We keep these languages
minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
in the example in Section~\ref{sec:intro} together with the type notation of
terms carried by the \elpiIn{lam} constructor.
%
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}

\noindent
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term \coqIn{P x} is represented as
\elpiIn{fapp[fuva N, x]}, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of \coqIn{P x} is instead \elpiIn{uva N [x]},
since unification variables come equipped with an explicit scope.
We say that the unification variable occurrence \elpiIn{uva N L} is in
\llambda if and only if \elpiIn{L} is made of distinct names. The
predicate to test this condition is called \elpiIn{pattern-fragment}:

@@

@@

\subsection{Equational theory and Unification}

In order to express properties \ref{prop:correct, prop:complete, prop:simulation}
we need to equip \Fo{} and \Ho with term equality,
substitution application and unification.

\paragraph{Term equality: \Eo vs. \Ee} We extend the equational theory
over ground terms to the full languages by adding the reflexivity of
unification variables (a variable is equal to itself).

>>>>

\begin{elpicode}
app[ lam A (a\ app [con"nfact", a, con"3"]) , x] =
app[ con"nfact"                             , N, NF]
\end{elpicode}
\noindent
~\\
The root cause of the problems we sketched in this example
is a subtle mismatch between the equational theories and
unification procedures of the meta language and the object language.

The equational theory of CIC is very rich. In
addition to the usual $\eta\beta$-equivalence for functions, terms (hence types)
are compared up to proposition unfolding and fixpoint unrolling. Still,
for efficiency and predictability reasons, most form of automatic proof search
employ a unification procedure that captures a simpler one,
just $\eta\beta$, and that solves higher-order problems
restricted to the pattern fragment \llambda.
We call this unification procedure \Uo{}.

In spite of the similarity the link between \Ue{} and \Uo{} is not trivial,
since the abstraction and application term constructors
the two unification procedures deal with are different. For example:\\
%
\setlength{\tabcolsep}{2pt}
\vspace{4pt}
\begin{tabular}{lcl|}
\elpiIn{x\ f x} & \Ue{} & \elpiIn{f}\\
\elpiIn{lam A x\ app[con"f", x]} & \Uo{} & \elpiIn{con"f"} \phantom{mn} \\
\elpiIn{lam A x\ app[con"f", x]} & \nUe{} & \elpiIn{con"f"}
\end{tabular}
\phantom{mn}
\begin{tabular}{lcl}
\elpiIn{P x} & \Ue{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \Uo{} & \elpiIn{x}\\
\elpiIn{app[P, x]} & \nUe{} & \elpiIn{x}\\
\end{tabular}
\vspace{4pt}
%
\paragraph{Contributions}
In this paper we identify a minimal language \Fo{} in which the problems
sketched in the introduction can be formally described.
We detail an encoding of a logic program on \Fo{} to a strongly related
logic program in \Ho (the language of the meta language) and we show that
the unification procedure of the meta language \Ue{} can be effectively
used to simulate a unification procedure \Uo for the object language that
features $\eta\beta$-conversion in the pattern-fragment.

\cref{sec:problem-statement} formally states the problem and gives the
intuition behind our solution. \cref{sec:encodings} discusses alternative
term encodings and related works. \cref{sec:grounwork} introduces
the languages \Fo{} and \Ho{}, \cref{sec:simulation} describes a
basic simulation of higher order logic programs.
\cref{sec:eta,sec:invariant1} completes its equational theory
with support for $\eta$-conversion. \cref{sec:beta} deals with the
practical necessity of ``tolerating'' terms outside of the
pattern-fragment and discusses how heuristic can be applied.
Finally \cref{sec:implementation} discusses the implementation in Elpi.

@@

\newcommand{\specunif}[3]{
  \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \exists \rho, %
      \rho #3_1 #1 \rho #3_2  %
        \Leftrightarrow #3_1 #2 #3_2 \mapsto \rho' \subseteq \rho
}

\newcommand{\unifcorrect}[3]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow
          \rho #3_1 #1 \rho #3_2  %
}

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\maybeeta}{\ensuremath{\Diamond\eta}\xspace}
\newcommand{\maybebeta}{\ensuremath{\Diamond\beta_0}\xspace}
\newcommand{\notllambda}{\ensuremath{\Diamond\llambda}\xspace}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

Even if we encountered the problem working on CIC we devise
a minimal setting to ease its study. In this setting we have
a \Fo{} language (for first order) with a rich(er) equational
theory and a \Ho{} meta language with a simpler one, and we reuse
the unification procedure of \Ho{} in order to implement one for \Fo.

\subsection{Preliminaries: \Fo{} and \Ho{}}

In order to reason about unification we provide a description of the
\Fo{} and \Ho languages where unification variables
are first class terms, i.e. they have a concrete syntax.
% We keep these languages
%minimal, for example, we omit the \elpiIn{all} quantifier of CIC we used
%in the example in Section~\ref{sec:intro} together with the type notation of
%terms carried by the \elpiIn{lam} constructor.
%
{
\setlength{\abovecaptionskip}{0pt}
\setlength{\belowcaptionskip}{-13pt}

\begin{figure}[H]
  \begin{tabular}{ll}
  \begin{minipage}{0.21\textwidth}
   \inputrawelpicode{code/fo_tm}
  \end{minipage}
  &
  \begin{minipage}{0.24\textwidth}
   \inputrawelpicode{code/ho_tm}
  \end{minipage}
  \end{tabular}\vspace{4pt}
  \caption{The \Fo{} and \Ho languages}\vspace{0.3em}
  \label{code:common-terms}
  \Description[code:common-terms]{code:common-terms}
\end{figure}
}
\noindent
Unification variables (\elpiIn{fuva} term constructor)
in \Fo{} have no explicit scope:
the arguments of an higher order variable are given via the \elpiIn{fapp}
constructor. For example the term <<\coqIn{P x}>> is represented as
<<\elpiIn{fapp[fuva N, x]}>>, where \elpiIn{N} is a memory address and
\elpiIn{x} is a bound variable.\\
In \Ho the representation of <<\coqIn{P x}>> is instead <<\elpiIn{uva N [x]}>>,
since unification variables are higher order and come equipped with an
explicit scope.
%\todo{use faddr for Fo}
%
%faddr and addr are memory addresses, the details are given in bla,
%here they are just unique identifiers for unif variables and their types
%do now allow to mistake one for the other.
%\todo{cleanup}

\paragraph{Notational conventions}

When we write \Ho terms outside code blocks we follow the
usual $\lambda$-calculus notation, reserving $f, g, a, b$ for constants,
$x, y, z$ for bound variables and $X, Y, Z, F, G, H$ for unification variables.
However we need to
distinguish between the ``application'' of a unification variable
to its scope and the application of a term to a list of arguments.
We write the scope of unification variables in subscript
while we use juxtaposition for regular application.
Here a few examples:\\
\vspace{4pt}
{
\setlength{\tabcolsep}{1em}
\begin{tabular}{ll}
  $f\appsep a$                  & \elpiIn{app [con "f", con "a"]}\\
  $\lambda x.\lambda y.F_{x y}$ & \elpiIn{lam x\ lam y\ uva F [x, y]} \\
  $\lambda x.F_{x} \appsep a$   & \elpiIn{lam x\ app [uva F [x], con "a"]} \\
  $\lambda x.F_{x} \appsep x$   & \elpiIn{lam x\ app [uva F [x], x]} \\
\end{tabular}
}
\vspace{4pt}

We use $s$, $s_1$, \ldots for terms in \Fo{} and $t$, $t_1$ \ldots for
terms in \Ho{}.

\subsection{Equational theories an unification}

In order to specify unification we need to
define the equational theory and
substitution (unification-variable assignment).

\subsubsection{Term equality: \Eo and \Ee}
For both languages we extend the equational theory
over ground terms to the full language by adding the reflexivity of
unification variables (a variable is equal to itself).

The first four rules are common to both equalities
and just define the usual congruence over terms. Since
we use an HOAS encoding they also capture $\alpha$-equivalence.
In addition to that \Eo has rules for $\eta$ and $\beta$-equivalence.

\noindent
The main point in showing these equality tests is to remark how
weaker \Ee is, and to identify the four rules that need special
treatment in the implementation of \Uo.
We defer the description of \elpiIn{beta} to \cref{sec:grounwork}, for now
it is sufficient to know that <<\elpiIn{beta F L R}>> computes in \elpiIn{R} the
weak head normal form of <<\elpiIn{app[F|L]}>>.

\paragraph{Substitution: $\rho s$ and $\sigma t$}

We write $\sigma = \{~ X \mapsto t ~\}$ the substitution that assigns
the term $t$ to the variable $X$.
We write $\sigma t$ for the application of
the substitution to a term $t$, and $\sigma X = \{~ \sigma t ~|~ t \in X ~\}$ when
$X$ is a set of terms.
We write $\sigma \subseteq \sigma'$ when $\sigma$ is more
general than $\sigma'$.
The domain of a substitution is the set of unification variables for which
it provides an assignment.
We write $\sigma \cup \sigma'$ set union to denote the concatenation of
two substitutions whose domains are disjoint.
We shall use $\rho$ for \Fo{} substitutions,
and $\sigma$ for the \Ho ones.
\todo{can be made a little more formal}
For brevity, in this section we consider
the substitution for \Fo{} and \Ho{} identical.
We defer to \cref{sec:grounwork} a more precise description
pointing out theirs differences.

Although we provide an implementation of \Ue in the supplementary material
(that we used for testing purposes) we only describe its signature here.

The meta language of choice is expected to provide 
an implementation of \Ue that satisfies
the following properties:
%~\ref{prop:correct-ml} and~\ref{prop:complete-ml}.
\begin{gather}
  \unifcorrect{\Ee}{\Ue}{t} \label{prop:correct-ml}\\
  \unifcomplete{\Ee}{\Ue}{t}\label{prop:complete-ml}
\end{gather}

%
% To state precisely the problem we solve we need a \Fo{} representation
% of CIC terms and a \Ho one. We extend the equality over ground
% terms to open ones with the reflexivity of unification variables (that is
% unification variable X is equal to itself, not to another one such a Y).
% We call \Eo that equality in \Fo{} and \Ee its counterpart in \Ho.
% We call \Uo the unification procedure we want to implement and
% \Ue the one provided by the meta language.

We write 
$\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ when
$\sigma t_1$ and $\sigma t_2$ unify with substitution $\sigma'$.
We write $t_1 \Ue t_2 \mapsto \sigma$ when
the initial substitution is empty.
Note that if $\sigma t_1 \Ue \sigma t_2 \mapsto \sigma'$ then
the domains of $\sigma$ and $\sigma'$ are disjoint.

Although we provide an implementation of \Uo{} in \cref{sec:founif},
our real goal is the simulation of an entire logic program.

\subsection{The problem: Logic Program Simulation}

We represent a logic program \emph{run} in \Fo{} as
a list \emph{steps} $p$ of length $\mathcal{N}$.
At each step $p$ we unify two terms $\foUnifPb_{p_l}$ and
$\foUnifPb_{p_r}$ taken from the set of all terms \foUnifPb.
\footnote{If the same rule is used multiple times in a run we
just consider as many copies as needed of the terms composing the
rules, with fresh unification variables each time.}
The composition of these steps starting from the
empty substitution $\rho_0$ produces the final
substitution $\rho_\mathcal{N}$, that is the result of the
logic-program execution.
%
\newcommand{\C}[4]{\ensuremath{\langle #1 \rangle}\mapsto(#2,#3,#4)}
\newcommand{\D}[4]{\ensuremath{\langle #1,#2,#3 \rangle^{-1}\mapsto #4}}
\newcommand{\progress}{\ensuremath{\mathrm{progress}}\xspace}
\newcommand{\fstep}{\ensuremath{\mathrm{fstep}}\xspace}
\newcommand{\hstep}{\ensuremath{\mathrm{hstep}}\xspace}
\newcommand{\frun}{\ensuremath{\mathrm{frun}}\xspace}
\newcommand{\hrun}{\ensuremath{\mathrm{hrun}}\xspace}
\newcommand{\stepF}[4]{\ensuremath{\fstep(#1,#2,#3) \mapsto #4}}
\newcommand{\stepFD}[5]{%
\ensuremath{#3 #1_{#2_l} \Uo #3 #1_{#2_r} \mapsto #4 \land #5 = #3 \cup #4}}
\newcommand{\stepH}[6]{\ensuremath{\hstep(#1,#2,#3,#4) \mapsto (#5, #6)}}
\newcommand{\stepHD}[6]{\ensuremath{%
#3 #1_{#2_l} \Ue #3 #1_{#2_r} \mapsto #4 \land \progress(#6,#3 \cup #4) \mapsto (#6',#5)}}
\newcommand{\runF}[3]{\ensuremath{\frun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runFD}[2]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepF{#1}{p}{\rho_{p-1}}{\rho_{p}}}}
\newcommand{\runH}[3]{\ensuremath{\hrun(#1,#2) \mapsto #3_{#2}}}
\newcommand{\runHD}[3]{\ensuremath{%
\bigwedge_{p = 1}^{#2} \stepH{#1}{p}{\sigma_{p-1}}{#3_{p-1}}{\sigma_{p}}{#3_p}}}
\newcommand{\deff}{\ensuremath{\stackrel{{\scriptscriptstyle def}}{=\!=\!\!\!=}}}
%
$$
\begin{array}{l}
\stepF{\foUnifPb}{p}{\rho}{\rho''}
\deff
\stepFD{\foUnifPb}{p}{\rho}{\rho'}{\rho''}\vspace{2pt}\\
\runF{\foUnifPb}{\mathcal{N}}{\rho}
\deff
\runFD{\foUnifPb}{\mathcal{N}}
\end{array}
$$

\noindent 
In order to simulate a \Fo{} logic program in \Ho{} we compile
each term $s \in \Fo{}$ into a term $t \in \Ho{}$. We write this
step $\C{s}{t}{m}{l}$. The implementation of the compiler
is detailed in \cref{sec:compilation,sec:eta,sec:beta}, here we just point
out that it additionally a variable mapping $m$ and list of links $l$.
The variable map connects unification variables in \Ho with variables
in \Fo{} and is used to ``decompile'' the assignment,
$\D{\sigma}{m}{l}{\rho}$. Links are an accessory piece of information whose
description is deferred to \cref{sec:nutshell}.

We simulate each run in \Fo{} with a run in \Ho as follows.
%Note that $\sigma_0$ is the empty substitution.
$$
\begin{array}{l}
\stepH{\hoUnifPb}{p}{\sigma}{\linkStore}{\sigma''}{\linkStore'} \deff\vspace{2pt}\\
  \qquad\stepHD{\hoUnifPb}{p}{\sigma}{\sigma'}{\sigma''}{\linkStore}\vspace{2pt}\\
\runH{\foUnifPb}{\mathcal{N}}{\rho} \deff \vspace{2pt}\\
  \qquad \hoUnifPb \times \mapStore \times \linkStore_0 = \{ (t_j,m_j,l_j) | s_j \in \foUnifPb, \C{s_j}{t_j}{m_j}{l_j} \}\vspace{2pt}\\
  \qquad \runHD{\hoUnifPb}{\mathcal{N}}{\linkStore}\vspace{2pt}\\
  \qquad \D{\sigma_{\mathcal{N}}}{\mapStore}{\linkStore_{\mathcal{N}}}{\rho_{\mathcal{N}}}
\end{array}
$$

\noindent
By analogy with \foUnifPb, we write $\hoUnifPb_{p_l}$ and $\hoUnifPb_{p_r}$
for the two \Ho{} terms being unified at step $p$, and we write $\hoUnifPb_p$
for the set $\{ \hoUnifPb_{p_l}, \hoUnifPb_{p_r} \}$.\\
\hstep{} is made of two sub-steps: a call to the meta language
unification and a check for \progress{} on the set of links, that intuitively
will compensate for the weaker equational theory honoured by \Ue.
\hrun{} compiles all terms in \foUnifPb{}, then executes each step and
finally decompiles the solution.
We claim:

@@

% We can define $s_1 \Uo{} s_2$% by specializing the code of \hrun{} to
% %$\foUnifPb = \{ s_1, s_2 \}$ 
% as follows:
% %
% $$
% \begin{array}{l}
% s_1 \Uo s_2 \mapsto \rho \deff \vspace{2pt}\\
% \quad\C{s_1}{t_1}{m_1}{l_1} \land \C{s_2}{t_2}{m_2}{l_2}\vspace{2pt}\\
% \quad    t_1 \Ue t_2 \mapsto \sigma' \land
%     \progress(\{l_1,l_2\},\sigma') \mapsto (L,\sigma'') \land\vspace{2pt}\\
% \quad \D{\sigma''}{\{m_1,m_2\}}{L}{\rho}
% \end{array}
% $$

% \begin{proposition}[Properties of \Uo{} in \llambda]
% \begin{gather}
%   \unifcorrect{\Eo}{\Uo}{s}\label{prop:correct}\\
%   \unifcomplete{\Eo}{\Uo}{s}\label{prop:complete}
% \end{gather}
% \end{proposition}

%In addition to correctness and completeness in \llambda,
We also claim that \hrun
handles terms outside \llambda in the following sense:

\begin{proposition}[Properties of \hrun outside \llambda]
\begin{gather}
  \exists \rho, \rho s_1 \Eo \rho s_2 \Rightarrow
  \C{s_i}{t_i}{m_i}{l_i} \Rightarrow
  t_1 \Ue t_2 \mapsto \sigma \label{prop:compilation-nllam}\\
  \sigma_{p-1} \hoUnifPb_{p} \in \llambda \Rightarrow \label{prop:step-nllam}\\
  \exists \rho,\stepF{\foUnifPb}{p}{\rho}{\rho'} \Leftrightarrow
  \stepH{\hoUnifPb}{p}{\sigma_{p-1}}{\linkStore_{p-1}}{\sigma_{p}}{\linkStore_p} \nonumber
\end{gather}
\end{proposition}

\noindent
Property \ref{prop:compilation-nllam} states that two terms for which
there is a unifier (given by an oracle and not necessarily a most general one),
then the compiler generates two terms that unify in \Ho.\\
%In other words the compiler moves out of the way all problematic terms.
Property \ref{prop:step-nllam} says that if the two terms involved in a step
re-enter \llambda, then \hstep succeeds. This is a typical example
in which the order of the unification problems in a logic-program run
does matter. The simplest example is the sequence $F \UnifRel \lambda x.a$ and
$F \appsep a \UnifRel a$: the second problem is not in \llambda and has two
unifiers, namely $\sigma_1 = \{~ F \mapsto \lambda x.x ~\}$ and
$\sigma_2 = \{~ F \mapsto \lambda x.a ~\}$. The first problem picks $\sigma_2$
making the second problem re-enter \llambda. In other words
\cref{prop:fidelity} \todo{...}

\subsection{The solution (in a nutshell)}
\label{sec:nutshell}
A term $s$ is compiled in a term $t$ where every
``problematic'' sub term $p$ is replaced by a fresh unification variable $h$
and an accessory \emph{link} that represents a suspended unification problem
$h \Ue p$. As a result \Ue is ``well behaved'' on $t$, in the sense that
it does not contradict \Eo as it would otherwise do on the
``problematic'' sub-terms. We now define ``problematic'' and ``well behaved''
more formally.

@@

\begin{definition}[\notllambda]\label{def:notllambda}
  $\notllambda = \{ X \appsep t_1 \ldots t_n ~|~ X \appsep t_1 \ldots t_n ~\mathrm{are~not~distinct~names} \}$.
\end{definition}

@@

% Note that proposition \ref{prop:nf} does not hold for \hrun as a whole
% since decompilation can introduce (actually restore) terms in
% \maybeeta, \maybebeta or \notllambda that were moved out of the way
% during compilation.

\section{Ground work for \Fo{}, \Ho and the compiler}
\label{sec:grounwork}
\todo{maybe move here napp beta deref fderef move}

We say that the unification variable occurrence \elpiIn{uva N L} is in
\llambda if and only if \elpiIn{L} is made of distinct names. The
predicate to test this condition is called \elpiIn{pattern-fragment}:

@@

%\input{code/move}

Applying the substitution corresponds to dereferencing a term with respect to
the memory.

To ease the comparison we split \Fo dereferencing into a
\elpiIn{fder} step and a \elpiIn{napp} one. The former step replaces references
to memory cells that are set with their values, and has a corresponding
operation in \Ho, namely \elpiIn{deref}. On the contrary \elpiIn{napp}
has no corresponding operation in \Ho. The reasons for this asymmetry is
that an \elpiIn{fapp} node with a flexible head is always mapped
to a \elpiIn{uva} (as per \cref{sec:compilation} and \cref{sec:beta}),
preventing nested applications to materialize.

\noindent
Note that move strongly relies on invariant \ref{inv:uvaarity}
that we shall define later.\todo{fixup}

@@

\todo{notation for problem compilation}
